---
title: "Reproductible Research - Peer Asssesment 1"
author: "Olivier Naeem"
date: "Sunday, July 20, 2014"
output:
  html_document:
    keep_md: yes
---

## Loading and preprocessing the data

### Configuration
Global configuration for the R code : 
``` {r global_options, echo=TRUE}
library(knitr)
library(xtable)
opts_chunk$set(echo= TRUE, results='asis')
```

### Creating the data frame
The file *activity.zip* is uncompressed in the **./data** directory. The **./data** directory is created if needed (this should happen only the first time).
The *activity.csv* file is then put in the **data** variable (a data frame) : 

``` {r loading_data}
# Create the data directory if not exists
if(!file.exists("./data")) {
  dir.create("./data")
}
# Uncompress the activity.csv file into the data directory 
unzip(zipfile = "./activity.zip", exdir = "./data", overwrite = T)

# Read the CSV file
data <- read.csv("./data/activity.csv")

#Displaying first values
xt <- xtable(data)
print(head(xt), type="html")
```


### Ordering the columns
I re-order the columns to my preference (date / interval / steps) : 

``` {r preprocessing}
data <- data[, c("date", "interval", "steps")]

#Displaying first values
xt <- xtable(data)
print(head(xt), type="html")
```

### Creating a new dataset

I aggregate all the data by date : 
``` {r aggregate}
data_by_day <- aggregate(steps ~ date, data = data, sum, na.rm=T)

#Displaying first values
xt <- xtable(data_by_day)
print(head(xt), type="html")

```

## What is mean total number of steps taken per day?

### Number of steps taken per day 
The total number of steps taken per day is figured out by the following histogram : 

``` {r steps_per_day}
barplot(data_by_day$steps, names.arg=data_by_day$date, main="Steps per day", xlab="Day", ylab="Nb of steps", col=rainbow(61))
```

### Mean and Median
I calculate now the mean and the median of the dataset : 

``` {r mean_and_median}
mean_by_day <- aggregate(steps ~ date, data = data, mean, na.rm=T)
median_by_day <- aggregate(steps ~ date, data = data, median, na.rm=T)
data_mean_median <- cbind(mean_by_day, median=median_by_day$steps)

#Displaying mean and median
xt <- xtable(data_mean_median)
print(xt, type="html")
```


``` {r plot_mean_and_median}
plot(mean_by_day$steps, main="Plot the averages steps per day", ylab="Nb of steps")

```

## What is the average daily activity pattern?

### Time series
I create a new data frame containing the average number of steps taken, averaged across all days : 
``` {r steps_and_interval}
data_avg_steps_by_5min <- aggregate(steps ~ interval, data = data, mean, na.rm=T)
#Displaying first values
xt <- xtable(data_avg_steps_by_5min)
print(head(xt), type="html")
```

The time series plot figuring out the daily activity is : 
``` {r plot_daily_activity}
plot(data_avg_steps_by_5min$steps ~ data_avg_steps_by_5min$interval, main="Daily activity", ylab = "Average nb of steps", xlab = "5-minute interval", type = "l")
```

### Analysing

The most active 5-min interval : 

``` {r most_active_interval}
five_min_interval <- data_avg_steps_by_5min[data_avg_steps_by_5min$steps == max(data_avg_steps_by_5min$steps), c("interval")]
print(five_min_interval)
```

## Imputing missing values

### Calculate and report the total number of missing value
Finding the number of missing values : 

``` {r missing_values}
nb_na_values <- colSums(is.na(data))

#Displaying number of NA values
print(nb_na_values, type="html")

```

### Devise a strategy for filling in all of the missing values in the dataset

I propose to replace the NA values with the mean values applying to this this specific  intervall.

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

``` {r dataset_with_no_missing_values}
# Calculate the mean for each 5-min interval
mean_by_interval <- aggregate(steps ~ interval, data = data, mean, na.rm=T)

data_no_na_values <- data
# Number of NA values
sum(is.na(data_no_na_values$steps))

for (i in 1:length(data$steps) ) {
  interval <- data$interval[i]
  if (is.na(data$steps[i])) 
      data_no_na_values$steps[i] <- mean_by_interval$steps[mean_by_interval$interval == interval]
}


# Remaining NA values : should be 0 !
sum(is.na(data_no_na_values$steps))

#Displaying first values
xt <- xtable(data_no_na_values)
print(head(xt), type="html")

```

### Make a histogram of the total number of steps taken each day

``` {r steps_per_day_no_na}
data_by_day_no_na <- aggregate(steps ~ date, data = data_no_na_values, sum, na.rm=T)

barplot(data_by_day_no_na$steps, names.arg=data_by_day_no_na$date, main="Steps per day", xlab="Day", ylab="Nb of steps", col=rainbow(61))
```

## Are there differences in activity patterns between weekdays and weekends?
